Software Architectures to Integrate Workflow Engines in Science Gateways

### 2016-04-26 paper.tex

\item Workflow integration: consists in adding a new workflow to the
  system so that users can execute it. It is triggered by an
  administrator of the science gateway and it results in an interface,
  for instance a web form, where users can enter the parameters of the
  workflow to be executed. The interaction has two aspects:
  (\texttt{a$_1$}) the programs used in the workflow are installed on
  the infrastructure, which may require administrative privileges on
  the infrastructure; (\texttt{a$_2$}) the workflow is configured in
  the science gateway so that it becomes available to users. Note that
  integrating a workflow is not the same process as integrating a
  workflow \emph{engine}.

\item Task control: operations to manage tasks on the infrastructure,
  including: authentication, submission, monitoring, termination,
  deletion, etc. Controlling tasks requires to deal with the
  heterogeneous batch managers and meta-schedulers that might be
  available on the infrastructure. When the infrastructure is a grid
  or a cloud, it may for instance be achieved using libraries that
  implement standards such as SAGA (Simple API for Grid
  Applications~\cite{goodale2006saga}), DRMAA (Distributed Resource
  Management Application API~\cite{troger2012distributed}), or OCCI
  (Open Cloud Computing Interface~\cite{edmonds2012toward}).
\item Data control: operations to manage data on the infrastructure,
  such as: upload, download, deletion, browsing, replication, caching,
  etc. Data movements can be triggered by the user in the science
  gateway (\texttt{$c_1$}), to upload input data or download processed
  data. They can also be performed by the workflow engine
  (\texttt{$c_2$}), to transfer workflow data (inputs, outputs, or
  intermediate results) across the infrastructure so that tasks can
  use it. The infrastructure might offer various data storage backends
  with heterogeneous interfaces. \closednote{Marc-Etienne}{this -c2- can be also trigerred
    by the user}{Tristan}{No, it can't: this is by definition an interaction between the workflow engine and the infrastructure. It is used to transfer data to/from tasks. I reworded a bit, is it clearer now?} Tools and services such as
  JSAGA~\cite{reynaud2010uniform} or Data Avenue~\cite{hajnal2014data}
  can be used to homogeneize these interfaces.
\item Workflow control: operations to control workflow execution in an
  engine, including: workflow submission, monitoring, termination,
  etc. Workflow control can be coarse-grained (black box) or
  fine-grained (white box). In a coarse-grained model, the various
  tasks created by a workflow execution are masked and the user only
  has a global view of the workflow execution. In a fine-grained
  model, user is exposed to the workflow topology, i.e. to the outputs
  of the individual tasks, their statuses and so on.
\item Sub-task control: operations used by tasks to submit sub-tasks
  on the infrastructure, including: submission, monitoring,
  termination, deletion, etc. Sub-task control is similar to
  interaction \texttt{b}, except that information about the parent
  task, which submitted a sub-task, is usually available and used for
  additional control. For instance, the parent task may wait for all
  its sub-tasks to complete before finishing, and conversely all the
  sub-tasks may be terminated if the parent task is killed.
\item Pool-agent: specific to the pool architecture described in
  Section~\ref{sec:pool}. This is an interaction used when agents
  retrieve work from a central pool. It covers agent registration and
  de-registration to the pool, protocols to send work from the pool to
  the agent, mechanisms to update work status, and so on. A similar
  type of interaction is used in pilot-job
  systems~\cite{turilli2015comprehensive} and other agent-based
  computing models.
\item Workflow conversion: translation from one workflow language to
  another one. This interaction may not be available or implementable
  for every workflow language. It has been developed mostly for
  well-structured and relatively simple workflow language such as between GWorkflowDL
  and Scufl~\cite{OLAB-09}, and between the 4 languages that were involved in the
  SHIWA initiative. In SHIWA, workflow conversion is done through an
  intermediate representation, the Interoperable Workflow Intermediate
  Representation~\cite{plankensteiner-montagnat-etal:2011}, which
  allows to convert among $n$ workflow languages using $2n$
  interactions instead of $n^2$.
\end{enumerate}

@@

\emph{Robustness of workflow execution} measures the likelihood that
workflow execution fails due to errors in the components or with the
interactions in the software architecture. Note that errors coming
from the infrastructure (e.g., unavailable data or terminated tasks) or
workflows (e.g., wrong user input or application errors) are not
covered since they do not stem directly from the software
architecture. Robustness is determined as the number of software
components and interactions that are specific to workflow execution,
which assumes that complex architectures tend to be prone to failure:
\begin{itemize}[leftmargin=0cm,itemindent=0.35cm,itemsep=0cm]
\item Components (\texttt{R$_1$}): number of specific components
  involved in workflow executions. Science
gateway, for instance, is \emph{not} specific to workflow execution
since it is used to authenticate users, add new workflows, transfer
user data, etc.
\item Interactions (\texttt{R$_2$}): number of specific interactions
  involved in workflow executions.
\end{itemize}
These specific components and interactions are represented in red in
the architecture diagrams in
Figure~\ref{fig:architectures}. \texttt{R$_1$} and \texttt{R$_2$}  are summed to obtain the total score for this
criterion, which measures the total number of software pieces that are
specifically involved in the workflow execution.

@@

\paragraph{Scalability} Adding a new engine instance requires a new
instance of the science gateway, which is in general not possible
(\texttt{S$_1$}=2)\note{Tristan}{What-about load-balancing between web
  server instances?}. Distributed engines are not available by default
(\texttt{S$_2$}=1). The scheduling of tasks on the infrastructure is
as complex as in any other architecture since the workflow engine
might implement any kind of scheduling policy (\texttt{S$_3$}=0).

@@

Regarding extensibility, most architectures are overall comparable,
except nested workflows (with service invocation) which is
significantly behind. This is explained by the complexity of the
nested workflow architecture, with 2 infrastructures and 2 workflow
services. Sub-tasking and workflow import (with service invocation)
perform a bit better when integrating new engines (\texttt{E$_1$}), which
is interesting given the current profusion of engines with different
characteristics.  However, sub-tasking, workflow import (with service
invocation) and tight integration perform worse than the others when adding
engine versions (\texttt{E$_2$}) due to the need to update
infrastructure (sub-tasking), science gateway (tight integration), or
workflow converter (workflow import). Workflow import performs worse
than the others when integrating new workflows (\texttt{E$_3$}) because of
the language conversion step. All architectures except nested
workflows perform the same when integrating new infrastructures (\texttt{E$_4$}).

The pool architecture is overall the most scalable, which is not a
surprise since it was designed precisely for scalability. Tight
integration is the least scalable and all the other architectures
perform the same. The overall scalability score, however, should not
mask the unique characteristics of architectures regarding this
criterion. Nested workflows are the only architecture that can easily
accommodate distributed workflow execution, which can be critical in
some cases. At the same time, the scheduling constraints created by
the sub-tasking and nested workflow architectures could well become
showstoppers depending on the type of the targeted
infrastructure. Non-reliable infrastructures, for example, could
hardly cope with workflow engines being wrapped in computing tasks as
done in sub-tasking. The availability of multiple engine instances
could also become a critical feature for science gateways with
important workloads, which would favor sub-tasking and pool.

@@

A few limitations remain that should be considered when using the
results of our evaluation. First of all, the evaluation is done at a
quite high-level of abstraction that may be a bit artificial when
applied to real systems. Abstracting architectures from their
implementation in specific systems is useful to compare them, but we
also realize that this comes at the cost of realism.  In particular,
the presented architectures are pure patterns that may be blended in
actual systems. The distinction between tight integration and service
invocation, for instance, may not be that clear in practice. Service
invocation may also be combined with sub-tasking in some cases.

@@

The particular case of interaction \texttt{g} (workflow language
conversion) should be further discussed since it is very likely that
this interaction could not be easily generalized to any workflow
language. For instance, converting FSL, PSOM or Nipype pipelines to
any other workflow language is very tricky because these engines rely
on general-purpose programming languages such as Bash, Octave and
Python, which are much richer than scientific workflow languages.

@@

Finally, the particular technical or historical context of a science
gateway project may obviously influence the choice of an architecture
to integrate workflow engines. For instance, some workflow engines may
be already available as web services, which would tend to favor
service invocation, and other science gateways may have strongly
tested task and data control features (interactions \texttt{b} and
\texttt{c}), which would favor sub-tasking. The migration cost between
architectures is totally ignored as well.

@@

We warmly thank Rafael Ferreira da Silva for its contribution to the
VIP platform and for creating Figure~\ref{fig:vip-architecture}. This
work is in the scope of the LABEX PRIMES (ANR-11- LABX-0063) of
Universit\'e de Lyon, within the program ``Investissements d’Avenir''
(ANR-11-IDEX-0007) operated by the French National Research Agency
(ANR). This work has been made possible with the support of the Irving
Ludmer Family Foundation and the Ludmer Centre for Neuroinformatics
and Mental Health.

>>>>

\item Workflow integration: consists in adding a new workflow to the
  system so that users can execute it. It is triggered by an
  administrator of the science gateway and it results in an interface,
  for instance a web form, where users can enter the parameters of the
  workflow to be executed. The interaction has two aspects:
  (\texttt{a$_1$}) the programs used in the workflow are installed on
  the infrastructure, which may require administrative privileges on
  the infrastructure; (\texttt{a$_2$}) the workflow is configured in
  the science gateway so that it becomes available to users. Note that
  integrating a workflow is not the same process as integrating a
  workflow \emph{engine}. \note{Marc-e}{automation?}
\item Task control: operations to manage tasks on the infrastructure,
  including: authentication, submission, monitoring, termination,
  deletion, etc. Controlling tasks requires to deal with the
  heterogeneous batch managers and meta-schedulers that might be
  available on the infrastructure. When the infrastructure is a grid
  or a cloud, it may for instance be achieved using libraries that
  implement standards such as SAGA (Simple API for Grid
  Applications~\cite{goodale2006saga}), DRMAA (Distributed Resource
  Management Application API~\cite{troger2012distributed}), or OCCI
  (Open Cloud Computing Interface~\cite{edmonds2012toward}).
\item Data control: operations to manage data on the infrastructure,
  such as: upload, download, deletion, browsing, replication, caching,
  etc. Data movements can be triggered by the user in the science
  gateway (\texttt{$c_1$}), to upload input data or download processed
  data. They can also be performed by the workflow engine
  (\texttt{$c_2$}), to transfer workflow data (inputs, outputs, or
  intermediate results) across the infrastructure so that tasks can
  use it. The infrastructure might offer various data storage backends
  with heterogeneous interfaces. \closenote{Marc-Etienne}{this -c2- can be also trigerred
    by the user}{Tristan}{No, it can't: this is by definition an interaction between the workflow engine and the infrastructure. It is used to transfer data to/from tasks. I reworded a bit, is it clearer now?}
    {Marc-e}{I meant, "using c1", a user in some platforms (CBRAIN), can trigger
    some operations across the infrastructure, as in c2... not just up/download data.
    And I clearly see now that this is in fact the case in III, c is not differentiated. Closing this thread for good.}
  Tools and services such as JSAGA~\cite{reynaud2010uniform} or Data Avenue~\cite{hajnal2014data}
  can be used to homogeneize these interfaces.
\item Workflow control: operations to control workflow execution in an
  engine, including: workflow submission, monitoring, termination,
  etc. Workflow control can be coarse-grained (black box) or
  fine-grained (white box). In a coarse-grained model, the various
  tasks created by a workflow execution are masked and the user only
  has a global view of the workflow execution. In a fine-grained
  model, user is exposed to the workflow topology, i.e. to the outputs
  of the individual tasks, their statuses and so on.
\item Sub-task control: operations used by tasks to submit sub-tasks
  on the infrastructure, including: submission, monitoring,
  termination, deletion, etc. Sub-task control is similar to
  interaction \texttt{b}, except that information about the parent
  task, which submitted a sub-task, is usually available and used for
  additional control. For instance, the parent task may wait for all
  its sub-tasks to complete before finishing, and conversely all the
  sub-tasks may be terminated if the parent task is killed.
\item Pool-agent: specific to the pool architecture described in
  Section~\ref{sec:pool}. This is an interaction used when agents
  retrieve work from a central pool. It covers agent registration and
  de-registration to the pool, protocols to send work from the pool to
  the agent, mechanisms to update work status, and so on. A similar
  type of interaction is used in pilot-job
  systems~\cite{turilli2015comprehensive} and other agent-based
  computing models.
\item Workflow conversion: translation from one workflow language to
  another one. This interaction may not be available or implementable
  for every workflow language. It has been developed mostly for
  well-structured and relatively simple workflow language such as between GWorkflowDL
  and Scufl~\cite{OLAB-09}, and between the 4 languages that were involved in the
  SHIWA initiative. In SHIWA, workflow conversion is done through an
  intermediate representation, the Interoperable Workflow Intermediate
  Representation~\cite{plankensteiner-montagnat-etal:2011}, which
  allows to convert among $n$ workflow languages using $2n$
  interactions instead of $n^2$.
\end{enumerate}

@@

\emph{Robustness of workflow execution} measures the likelihood that
workflow execution fails due to errors in the components or with the
interactions in the software architecture. Note that errors coming
from the infrastructure (e.g., unavailable data or terminated tasks) or
workflows (e.g., wrong user input or application errors) are not
covered since they do not stem directly from the software
architecture. Robustness is determined as the number of software
components and interactions that are specific to workflow execution,
which assumes that complex architectures tend to be more prone to failure:
\begin{itemize}[leftmargin=0cm,itemindent=0.35cm,itemsep=0cm]
\item Components (\texttt{R$_1$}): number of specific components
  involved in workflow executions. Science
gateway, for instance, is \emph{not} specific to workflow execution
since it is used to authenticate users, add new workflows, transfer
user data, etc.
\item Interactions (\texttt{R$_2$}): number of specific interactions
  involved in workflow executions.
\end{itemize}
These specific components and interactions are represented in red in
the architecture diagrams in
Figure~\ref{fig:architectures}. \texttt{R$_1$} and \texttt{R$_2$}  are summed to obtain the total score for this
criterion, which measures the total number of software pieces that are
specifically involved in the workflow execution.

@@

\paragraph{Scalability} Adding a new engine instance requires a new
instance of the science gateway, which is in general not possible
(\texttt{S$_1$}=2)\note{Tristan}{What-about load-balancing between web
  server instances?} \note{Marc-e}{Would that really be considered a real architectural feature or an IT hack?}.
  Distributed engines are not available by default
(\texttt{S$_2$}=1). The scheduling of tasks on the infrastructure is
as complex as in any other architecture since the workflow engine
might implement any kind of scheduling policy (\texttt{S$_3$}=0).

@@

Regarding extensibility, most architectures are overall comparable,
except nested workflows (with service invocation) which is
significantly behind. This is explained by the complexity of the
nested workflow architecture, with 2 infrastructures and 2 workflow
services. Sub-tasking and workflow import (with service invocation)
perform a bit better when integrating new engines (\texttt{E$_1$}), which
is interesting given the current profusion of engines with different
characteristics.  However, sub-tasking, workflow import (with service
invocation) and tight integration perform worse than the others when adding
engine versions (\texttt{E$_2$}) due to the need to update
infrastructure \note{Marc-e}{confusion with physical?}(sub-tasking), science gateway (tight integration), or
workflow converter (workflow import). Workflow import performs worse
than the others when integrating new workflows (\texttt{E$_3$}) because of
the language conversion step. All architectures except nested
workflows perform the same when integrating new infrastructures (\texttt{E$_4$}).

The pool architecture is overall the most scalable, which is not a
surprise since it was designed precisely for scalability. Tight
integration is the least scalable and all the other architectures
perform the same. The overall scalability score, however, should not
mask the unique characteristics of architectures regarding this
criterion. Nested workflows are the only architecture that can easily
accommodate distributed workflow execution, which can be critical in
some cases. At the same time, the scheduling constraints created by
the sub-tasking and nested workflow architectures could well become
showstoppers depending on the type of the targeted
infrastructure. Non-reliable infrastructures, for example, could
hardly cope with workflow engines being wrapped in computing tasks as
done in sub-tasking \note{Marc-e}{free for all? WE dev involved?}. The availability of multiple engine instances
could also become a critical feature for science gateways with
important workloads, which would favor sub-tasking and pool.

@@

A few limitations remain that should be considered when using the
results of our evaluation. First of all, the evaluation is done at a
quite high-level of abstraction that may be a bit artificial when
applied to real systems. Abstracting architectures from their
implementation in specific systems is useful to compare them, but we
also realize that this comes at the cost of realism. In particular,
the presented architectures are theoritical patterns that may be blended in
actual systems. The distinction between tight integration and service
invocation, for instance, may not be always that clear in practice. Service
invocation may also be combined with sub-tasking in some cases. However, the
criteria discussed in this work would still apply to broadly compare and categorize
such cases of hydrid architectures.

@@

The particular case of interaction \texttt{g} (workflow language
conversion) should be further discussed since it is very likely that
this interaction could not be easily generalized to any workflow
language. For instance, converting FSL, PSOM or Nipype pipelines to
any other workflow language is problematic because these engines rely
on general-purpose programming languages such as Bash, Octave and
Python, which are much richer than scientific workflow languages.

@@

Finally, the particular technical or historical context of a science
gateway project may obviously influence the choice of an architecture
to integrate workflow engines. For instance, some workflow engines may
be already available as web services, which would tend to favor
service invocation, and other science gateways may have strongly
tested task and data control features (interactions \texttt{b} and
\texttt{c}), which would favor sub-tasking. The migration cost between
architectures has been ignored as well.

@@

We warmly thank Rafael Ferreira da Silva for its contribution to the
VIP platform and for creating Figure~\ref{fig:vip-architecture}. This
work is in the scope of the LABEX PRIMES (ANR-11- LABX-0063) of
Universit\'e de Lyon, within the program ``Investissements d’Avenir''
(ANR-11-IDEX-0007) operated by the French National Research Agency
(ANR). This work has been made possible with the support of the Irving
Ludmer Family Foundation and the Ludmer Centre for Neuroinformatics
and Mental Health. \note{Marc-e}{Would it be proper to thank EGI? Compute Canada? CANARIE?}
